---
title: "data cleaning"
output: html_document
date: "2024-10-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# packages required for this script
library(readr)
library(dplyr)
library(lubridate)
library(tidyr)
library(stringr)
```

```{r import data}
################################################################################
##################   Import CPET data   ########################################
################################################################################
folder_path <- "PANAMA_CPET_files_download"

# importing all the CPET files from the folder

files <- list.files(path = folder_path, pattern = "\\.TXT$", full.names = TRUE)
all_data <- list()
pan_id_counter <- list()

# Looping through each file
for (file in files) {
  file_name <- basename(file)
  pan_id <- sub("^(PAN_\\d{4}).*", "\\1", file_name)
  data <- read_delim(file, delim = "!", escape_double = FALSE,  locale = locale(decimal_mark = ","), trim_ws = TRUE)
  
  # Account for multiple tests for 1 participant
  if (!is.null(pan_id_counter[[pan_id]])) {
    pan_id_counter[[pan_id]] <- pan_id_counter[[pan_id]] + 1
  } else {
    pan_id_counter[[pan_id]] <- 1
  }
  data$PAN_ID <- paste0(pan_id, "_", pan_id_counter[[pan_id]])
  
  all_data[[file_name]] <- data
}

cpet_data <- do.call(rbind, all_data)

# Rename variables with spaces
cpet_data <- cpet_data %>%
  rename(bp_sys = "BP sys", bp_dia = "BP dia")

# Remove rows where time is NA
cpet_data <- cpet_data %>%
  filter(!is.na(Time))

# Reformat time variable 
seconds <- as.numeric(cpet_data$Time/60)
cpet_data$Time <- hms::as_hms(seconds)

cpet_data$PAN_ID_short <- str_extract(cpet_data$PAN_ID, "^PAN_\\d{4}")

# Keep variables of interest
cpet_data <- cpet_data[,c("PAN_ID","PAN_ID_short","Time","Step","Load","VE","HR","RR","VO2","VCO2")]

# # Export, to have the csv at hand
# write.csv(cpet_data, file = "cpet_data.csv", row.names = FALSE)
# # Import file
# cpet_data <- read_csv("cpet_data.csv")

################################################################################
################   Import REDCap data (descriptives)  ##########################
################################################################################

redcap_data <- read_csv("panama_redcap.csv")

## Join the clinical and descriptives into one row, no need to have them separated - one row per participant
redcap_data <- redcap_data %>%
  group_by(record_id) %>%
  summarise(across(everything(), ~ first(na.omit(.)), .names = "{col}")) %>%
  ungroup()

redcap_data <- redcap_data %>%
  rename(PAN_ID = cpet_id)
redcap_data$PAN_ID <- as.character(redcap_data$PAN_ID)

# add PAN_ID of participant with empty ID field (PAN_0141)
redcap_data <- redcap_data %>%
  mutate(PAN_ID = ifelse(record_id == 141, "PAN_0141", PAN_ID))

# Correct structure of the variables
redcap_data <- redcap_data %>%
  mutate(
    ethnic = as.factor(ethnic),
    smoke = as.factor(smoke),
    sex = as.factor(sex),
    disease = as.factor(disease),
    cpet_protocol = as.factor(cpet_protocol),
    copd_sev = as.factor(copd_sev),
    asthma_sev = as.factor(asthma_sev)
  )

# Rename IDs with spelling errors
redcap_data$PAN_ID[redcap_data$PAN_ID=='pAN_0043'] <- 'PAN_0043'
redcap_data$PAN_ID[redcap_data$PAN_ID=='PAN_110'] <- 'PAN_0110'
redcap_data$PAN_ID[redcap_data$PAN_ID=='PAN_154'] <- 'PAN_0154'
redcap_data$PAN_ID[redcap_data$PAN_ID=='PAN_159'] <- 'PAN_0159'
redcap_data$PAN_ID[redcap_data$PAN_ID=='PAN_161'] <- 'PAN_0161'
redcap_data$PAN_ID[redcap_data$PAN_ID=='PAN_162'] <- 'PAN_0162'
redcap_data$PAN_ID[redcap_data$PAN_ID=='PAN_164'] <- 'PAN_0164'
redcap_data$PAN_ID[redcap_data$PAN_ID=='PAN_172'] <- 'PAN_0172'


# Delete unnecessary variables - need to rewrite
redcap_data$record_id <- NULL
redcap_data$redcap_repeat_instance <-NULL
redcap_data$redcap_repeat_instrument <- NULL
redcap_data$nhc <- NULL
redcap_data$identification_complete <- NULL
redcap_data$clinical_data_complete <- NULL
redcap_data$cpet_field <- NULL
redcap_data$reason_stop <- NULL
redcap_data$dyspnea_pre <- NULL
redcap_data$fatigue_pre <- NULL
redcap_data$dyspnea_post <- NULL
redcap_data$fatigue_post <- NULL
redcap_data$treatment___1 <- NULL
redcap_data$treatment___2 <- NULL
redcap_data$treatment___3 <- NULL
redcap_data$treatment___4 <- NULL
redcap_data$treatment___5 <- NULL
redcap_data$treatment___6 <- NULL
redcap_data$treatment___7 <- NULL
redcap_data$treatment___8 <- NULL
redcap_data$o2_dose <- NULL
redcap_data$o2_freq <- NULL
redcap_data$cv_treat___1 <- NULL
redcap_data$cv_treat___2 <- NULL
redcap_data$cv_treat___3 <- NULL
redcap_data$cv_treat___4 <- NULL
redcap_data$other_treat <- NULL
redcap_data$comments <- NULL

# # save to have csv in folder
# write.csv(redcap_data,"redcap_data.csv",row.names = FALSE)
# # import
# redcap_data <- read.csv("redcap_data.csv")

# join descriptives and CPET data
PAN_data <- cpet_data %>%
  left_join(redcap_data, by = c("PAN_ID_short" = "PAN_ID"))


```

CLEAN DATA

Remove recovery time (as this could affect the prediction)
```{r}
PAN_data <- PAN_data %>%
  filter(!(Step %in% c("Rep", "R 2", "R 1")))
```

Missings
```{r}
# Table with NAs in each variable
missing_summary <- PAN_data %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Missing_Count")

print(missing_summary)
```

TIME RESOLUTIONS - 5s, 15s, 30s, 1min 
```{r}
# currently at breath by breath

# Time in POSIXct format
cpet_data$Time <- as.POSIXct(cpet_data$Time, format = )

###### 5 second intervals
cpet_data_10s <- cpet_data %>%
  mutate(Time = floor_date(Time, unit = "1 second")) %>% 
  group_by(PAN_ID, Time_10s) %>%
  summarise(across(everything(), mean, na.rm = TRUE), .groups = 'drop')

cpet_data_10s$Time_10s <- as.numeric(cpet_data_10s$Time_10s)
# Fix Time variable
seconds <- as.numeric(cpet_data_10s$Time_10s)
cpet_data_10s$Time_10s <- hms::as_hms(seconds)
cpet_data_10s$Time <- cpet_data_10s$Time_10s


######  30 second intervals
cpet_data_30s <- cpet_data %>%
  mutate(Time_30s = floor_date(Time, unit = "30 seconds")) %>%
  group_by(PAN_ID, Time_30s) %>%
  summarise(across(everything(), mean, na.rm = TRUE), .groups = 'drop')

cpet_data_30s$Time_30s <- as.numeric(cpet_data_30s$Time_30s)
# Fix Time variable
seconds <- as.numeric(cpet_data_30s$Time_30s)
cpet_data_30s$Time_30s <- hms::as_hms(seconds)
cpet_data_30s$Time <- cpet_data_30s$Time_30s


######  1 minute intervals
cpet_data_60s <- cpet_data %>%
  mutate(Time_60s = floor_date(Time, unit = "1 minute")) %>%
  group_by(PAN_ID, Time_60s) %>%
  summarise(across(everything(), mean, na.rm = TRUE), .groups = 'drop')
cpet_data_60s$Time_60s <- as.numeric(cpet_data_60s$Time_60s)
# Fix Time variable
seconds <- as.numeric(cpet_data_60s$Time_60s)
cpet_data_60s$Time_60s <- hms::as_hms(seconds)
cpet_data_60s$Time <- cpet_data_60s$Time_60s

# merge redcap data with aggregated CPET data
PAN_data <- cpet_data %>%
  left_join(redcap_data, by = "PAN_ID")

PAN_data_10s <- cpet_data_10s %>%
  left_join(redcap_data, by = "PAN_ID")

PAN_data_30s <- cpet_data_30s %>%
  left_join(redcap_data, by = "PAN_ID")

PAN_data_60s <- cpet_data_60s %>%
  left_join(redcap_data, by = "PAN_ID")

```

GENERATE VARIABLES
```{r}
# BMI (weight in kg/ height in m2)
PAN_data$bmi <- PAN_data$weight / ((PAN_data$height/100)^2)

# METs
PAN_data$METs <- ((PAN_data$VO2 *1000) / PAN_data$weight ) / 3.5

# Respiratory Exchange Ratio
PAN_data$RER <- PAN_data$VCO2 / PAN_data$VO2

# VE max
PAN_data <- PAN_data %>%
  group_by(PAN_ID) %>%
  mutate(VE_max = max(VE, na.rm = TRUE)) %>%
  ungroup()

# RR max
PAN_data <- PAN_data %>%
  group_by(PAN_ID) %>%
  mutate(RR_max = max(RR, na.rm = TRUE)) %>%
  ungroup()

# HR max
PAN_data <- PAN_data %>%
  group_by(PAN_ID) %>%
  mutate(HR_max = max(HR, na.rm = TRUE)) %>%
  ungroup()

# VO2 max
PAN_data <- PAN_data %>%
  group_by(PAN_ID) %>%
  mutate(VO2_max = max(VO2, na.rm = TRUE)) %>%
  ungroup()
```

Remove outliers (by participant) for time series variables (VE, HR and RR)
1. Group by participant
2. calculate moving average
3. calculate moving sd
4. identify values that are outside of 3 SDs from the mean
5. remove them

```{r remove outliers}
library(dplyr)
library(zoo)

window_size <- 30 #have tried other values but 30 is the most backed by the literature

# minute ventilation 
PAN_data <- PAN_data %>%
  group_by(PAN_ID) %>%  # calculate mean and sd by participant
  mutate(
    MA = rollapply(VE, width = window_size, FUN = mean, align = "center", fill = NA),
    SD = rollapply(VE, width = window_size, FUN = sd, align = "center", fill = NA)
  ) %>%
  mutate(
    outlier = abs(VE - MA) > 3 * SD # detect datapoints outside of 3 SDs from the local mean
  ) %>%
  filter(!outlier) %>% # remove outlier datapoints
  ungroup()

# heart rate
PAN_data <- PAN_data %>%
  group_by(PAN_ID) %>% 
  mutate(
    MA = rollapply(HR, width = window_size, FUN = mean, align = "center", fill = NA),
    SD = rollapply(HR, width = window_size, FUN = sd, align = "center", fill = NA)
  ) %>%
  mutate(
    outlier = abs(HR - MA) > 3 * SD
  ) %>%
  filter(!outlier) %>%
  ungroup()

# breathing rate (outliers as NA)
PAN_data <- PAN_data %>%
  group_by(PAN_ID) %>%
  mutate(
    MA = rollapply(RR, width = window_size, FUN = mean, align = "center", fill = NA),
    SD = rollapply(RR, width = window_size, FUN = sd, align = "center", fill = NA),
    outlier = abs(RR - MA) > 3 * SD,
    RR = if_else(outlier, NA_real_, RR)
  ) %>%
  ungroup()

# breathing rate (removing outlier rows) - for the model
PAN_data <- PAN_data %>%
  group_by(PAN_ID) %>% 
  mutate(
    MA = rollapply(RR, width = window_size, FUN = mean, align = "center", fill = NA),
    SD = rollapply(RR, width = window_size, FUN = sd, align = "center", fill = NA)
  ) %>%
  mutate(
    outlier = abs(RR - MA) > 3 * SD
  ) %>%
  filter(!outlier) %>%
  ungroup()
```

Delete datapoints past anaerobic threshold (RER >= 1.0 consistently for 15 seconds)
```{r}
anaerobic_threshold <- PAN_data %>%
  group_by(PAN_ID) %>%
  arrange(Time) %>%
  mutate(RER_above_1 = RER >= 1.0,                  # Identify where RER > 1.0
         anaerobic_threshold = cumsum(!RER_above_1),    # Reset when RER <= 1.0
         flag = ave(RER_above_1, anaerobic_threshold, 
                    FUN = cumsum) > 15) %>%
  ungroup() %>%
  filter(flag)

PAN_data <- PAN_data %>%
  anti_join(anaerobic_threshold, by = c("PAN_ID", "Time"))
```

Disease as a factor variable for the cohort separation
```{r}
PAN_data <- PAN_data %>%
  mutate(
    disease = factor(
      disease,
      levels = c(1, 2, 3),
      labels = c("Healthy", "COPD", "Asthma")
    )
  )
```

```{r datasets by cohorts}
data_healthy <- PAN_data[PAN_data$disease == "Healthy",]
data_COPD <- PAN_data[PAN_data$disease == "COPD",]
data_asthma <- PAN_data[PAN_data$disease == "Asthma",]
```

Exercise intensity calculations
```{r EI healthy}
# by age
data_healthy <- data_healthy %>%
  mutate(
    ex_intensity = case_when(
      # Age 20-39
      age <= 39 & METs < 2.4 ~ "Very Light",
      age <= 39 & METs >= 2.4 & METs <= 4.7 ~ "Light",
      age <= 39 & METs >= 4.8 & METs <= 7.1 ~ "Moderate",
      age <= 39 & METs >= 7.2 & METs <= 10.1 ~ "Vigorous",
      age <= 39 & METs >= 10.2 ~ "Near to Maximal",
      
      # Age 40-64
      age >= 40 & age <= 64 & METs < 2.0 ~ "Very Light",
      age >= 40 & age <= 64 & METs >= 2.0 & METs <= 3.9 ~ "Light",
      age >= 40 & age <= 64 & METs >= 4.0 & METs <= 5.9 ~ "Moderate",
      age >= 40 & age <= 64 & METs >= 6.0 & METs <= 8.4 ~ "Vigorous",
      age >= 40 & age <= 64 & METs >= 8.5 ~ "Near to Maximal",
      
      # Age >= 65
      age >= 65 & METs < 1.6 ~ "Very Light",
      age >= 65 & METs >= 1.6 & METs <= 3.1 ~ "Light",
      age >= 65 & METs >= 3.2 & METs <= 4.7 ~ "Moderate",
      age >= 65 & METs >= 4.8 & METs <= 6.7 ~ "Vigorous",
      age >= 65 & METs >= 6.8 ~ "Near to Maximal",
      
      # Default case for missing values or undefined conditions
      TRUE ~ NA_character_
    )
  )
data_healthy <- data_healthy %>%
  mutate(
    ex_intensity = factor(
      ex_intensity,
      levels = c("Very Light", "Light", "Moderate", "Vigorous", "Near to Maximal"),
      ordered = TRUE
    )
  )

# Create a bar plot to see the distribution
ggplot(data_healthy, aes(x = ex_intensity)) +
  geom_bar(fill = "skyblue", color = "black") +
  labs(
    title = "Healthy distribution of Exercise Intensity using METs by age",
    x = "Exercise Intensity",
    y = "Count"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
```


```{r by ages on COPD}
data_COPD <- data_COPD %>%
  mutate(
    ex_intensity = case_when(
      # Age 20-39
      age <= 39 & METs < 2.4 ~ "Very Light",
      age <= 39 & METs >= 2.4 & METs <= 4.7 ~ "Light",
      age <= 39 & METs >= 4.8 & METs <= 7.1 ~ "Moderate",
      age <= 39 & METs >= 7.2 & METs <= 10.1 ~ "Vigorous",
      age <= 39 & METs >= 10.2 ~ "Near to Maximal",
      
      # Age 40-49
      age >= 40 & age <= 49 & METs < 2.0 ~ "Very Light",
      age >= 40 & age <= 49 & METs >= 2.0 & METs <= 3.9 ~ "Light",
      age >= 40 & age <= 49 & METs >= 4.0 & METs <= 5.9 ~ "Moderate",
      age >= 40 & age <= 49 & METs >= 6.0 & METs <= 8.4 ~ "Vigorous",
      age >= 40 & age <= 49 & METs >= 8.5 ~ "Near to Maximal",
      
      # Age >= 50 (COPD patients over 50 are considered older adults (ACSM))
      age >= 50 & METs < 1.6 ~ "Very Light",
      age >= 50 & METs >= 1.6 & METs <= 3.1 ~ "Light",
      age >= 50 & METs >= 3.2 & METs <= 4.7 ~ "Moderate",
      age >= 50 & METs >= 4.8 & METs <= 6.7 ~ "Vigorous",
      age >= 50 & METs >= 6.8 ~ "Near to Maximal",
      
      TRUE ~ NA_character_
    )
  )

data_COPD <- data_COPD %>%
  mutate(
    ex_intensity = factor(
      ex_intensity,
      levels = c("Very Light", "Light", "Moderate", "Vigorous", "Near to Maximal"),
      ordered = TRUE
    )
  )


# Create a bar plot to see the distribution
ggplot(data_COPD, aes(x = ex_intensity)) +
  geom_bar(fill = "skyblue", color = "black") +
  labs(
    title = "COPD Distribution of Exercise Intensity using METS and age",
    x = "Exercise Intensity",
    y = "Count"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```



EXPLORATORY DATA ANALYSIS
using Datamaid
```{r}
library(dataMaid)
# generate reports
makeDataReport(PAN_data, output = "html", replace = TRUE)
makeDataReport(data_COPD, output = "html", replace = TRUE)
makeDataReport(data_healthy, output = "html", replace = TRUE)
makeDataReport(data_asthma, output = "html", replace = TRUE)

```


DESCRIPTIVES

```{r}
### PARTICIPANT DESCRIPTORS

library(table1)

table1(~ factor(sex) + age + height + weight + factor(ethnic) +  factor(smoke) + factor(asthma_sev) + factor(copd_sev) | disease, data=redcap_data, overall=c(left="Total"))

# treatment table
table1(~ factor(treatment___1) + factor(treatment___2) + factor(treatment___3) + factor(treatment___4) + factor(treatment___5) + factor(treatment___6) + factor(treatment___7) + factor(treatment___8)| disease, data=redcap_data, overall=c(left="Total"))

```


Descriptives of physiological variables

```{r}

## PHYSIOLOGICAL VARIABLES TABLE DESCRIPTORS (missing some, i dont know what they are called in the ds)
table1(~ VE + Ti + RR + VO2 + HR | disease, data=PAN_data, overall=c(left="Total"))

# Rest is when step is Rep (will only include some of the participants as they dont all have rep)
data_rest <- PAN_data[(PAN_data$Step=='Rep'),]

# table for reposo dataset
table1(~ VE + Ti + RR + VO2 + HR | disease, data=data_rest, overall=c(left="Total"))

# FEV1 , FVC , etc
table1(~ fev1 + fev1_pred + fvc + fvc_pred + fev1_fvc_r | disease, data = redcap_data, overall = c(left = "Total"))
```


Maximum values' descriptives
```{r}
### VE max mean(SD)
# Total
summary_stats <- PAN_data %>%
  group_by(PAN_ID) %>%
  summarise(VE_max = unique(VE_max)) %>% # Take unique VE_max per participant
  summarise(mean_VE_max = mean(VE_max, na.rm = TRUE),
            sd_VE_max = sd(VE_max, na.rm = TRUE))
summary_stats
# No RD
summary_stats <- data_healthy %>%
  group_by(PAN_ID) %>%
  summarise(VE_max = unique(VE_max)) %>% # Take unique VE_max per participant
  summarise(mean_VE_max = mean(VE_max, na.rm = TRUE),
            sd_VE_max = sd(VE_max, na.rm = TRUE))
summary_stats
# COPD
summary_stats <- data_COPD %>%
  group_by(PAN_ID) %>%
  summarise(VE_max = unique(VE_max)) %>% # Take unique VE_max per participant
  summarise(mean_VE_max = mean(VE_max, na.rm = TRUE),
            sd_VE_max = sd(VE_max, na.rm = TRUE))
summary_stats
# Asthma
summary_stats <- data_asthma %>%
  group_by(PAN_ID) %>%
  summarise(VE_max = unique(VE_max)) %>% # Take unique VE_max per participant
  summarise(mean_VE_max = mean(VE_max, na.rm = TRUE),
            sd_VE_max = sd(VE_max, na.rm = TRUE))
summary_stats

### RR max mean(SD)
# Total
summary_stats <- PAN_data %>%
  group_by(PAN_ID) %>%
  summarise(RR_max = unique(RR_max)) %>% # Take unique RR_max per participant
  summarise(mean_RR_max = mean(RR_max, na.rm = TRUE),
            sd_RR_max = sd(RR_max, na.rm = TRUE))
summary_stats
# No RD
summary_stats <- data_healthy %>%
  group_by(PAN_ID) %>%
  summarise(RR_max = unique(RR_max)) %>% # Take unique RR_max per participant
  summarise(mean_RR_max = mean(RR_max, na.rm = TRUE),
            sd_RR_max = sd(RR_max, na.rm = TRUE))
summary_stats
# COPD
summary_stats <- data_COPD %>%
  group_by(PAN_ID) %>%
  summarise(RR_max = unique(RR_max)) %>% # Take unique RR_max per participant
  summarise(mean_RR_max = mean(RR_max, na.rm = TRUE),
            sd_RR_max = sd(RR_max, na.rm = TRUE))
summary_stats
# Asthma
summary_stats <- data_asthma %>%
  group_by(PAN_ID) %>%
  summarise(RR_max = unique(RR_max)) %>% # Take unique RR_max per participant
  summarise(mean_RR_max = mean(RR_max, na.rm = TRUE),
            sd_RR_max = sd(RR_max, na.rm = TRUE))
summary_stats

### HR max mean(SD)
# Total
summary_stats <- PAN_data %>%
  group_by(PAN_ID) %>%
  summarise(HR_max = unique(HR_max)) %>% # Take unique HR_max per participant
  summarise(mean_HR_max = mean(HR_max, na.rm = TRUE),
            sd_HR_max = sd(HR_max, na.rm = TRUE))
summary_stats
# No RD
summary_stats <- data_healthy %>%
  group_by(PAN_ID) %>%
  summarise(HR_max = unique(HR_max)) %>% # Take unique HR_max per participant
  summarise(mean_HR_max = mean(HR_max, na.rm = TRUE),
            sd_HR_max = sd(HR_max, na.rm = TRUE))
summary_stats
# COPD
summary_stats <- data_COPD %>%
  group_by(PAN_ID) %>%
  summarise(HR_max = unique(HR_max)) %>% # Take unique HR_max per participant
  summarise(mean_HR_max = mean(HR_max, na.rm = TRUE),
            sd_HR_max = sd(HR_max, na.rm = TRUE))
summary_stats
# Asthma
summary_stats <- data_asthma %>%
  group_by(PAN_ID) %>%
  summarise(HR_max = unique(HR_max)) %>% # Take unique HR_max per participant
  summarise(mean_HR_max = mean(HR_max, na.rm = TRUE),
            sd_HR_max = sd(HR_max, na.rm = TRUE))
summary_stats

### VO2 max mean(SD)
# Total
summary_stats <- PAN_data %>%
  group_by(PAN_ID) %>%
  summarise(VO2_max = unique(VO2_max)) %>% # Take unique VO2_max per participant
  summarise(mean_VO2_max = mean(VO2_max, na.rm = TRUE),
            sd_VO2_max = sd(VO2_max, na.rm = TRUE))
summary_stats
# No RD
summary_stats <- data_healthy %>%
  group_by(PAN_ID) %>%
  summarise(VO2_max = unique(VO2_max)) %>% # Take unique VO2_max per participant
  summarise(mean_VO2_max = mean(VO2_max, na.rm = TRUE),
            sd_VO2_max = sd(VO2_max, na.rm = TRUE))
summary_stats
# COPD
summary_stats <- data_COPD %>%
  group_by(PAN_ID) %>%
  summarise(VO2_max = unique(VO2_max)) %>% # Take unique VO2_max per participant
  summarise(mean_VO2_max = mean(VO2_max, na.rm = TRUE),
            sd_VO2_max = sd(VO2_max, na.rm = TRUE))
summary_stats
# Asthma
summary_stats <- data_asthma %>%
  group_by(PAN_ID) %>%
  summarise(VO2_max = unique(VO2_max)) %>% # Take unique VO2_max per participant
  summarise(mean_VO2_max = mean(VO2_max, na.rm = TRUE),
            sd_VO2_max = sd(VO2_max, na.rm = TRUE))
summary_stats
```

Length of CPETs
```{r}
## healthy cohort
test_lengths <- data_healthy %>%
  group_by(PAN_ID) %>%
  summarise(test_length = max(Time, na.rm = TRUE)) %>%
  ungroup()

test_lengths <- test_lengths %>%
  mutate(test_length = as.numeric(test_length),
         test_length_rounded = round(test_length / 60))

test_length_counts <- test_lengths %>%
  count(test_length_rounded, name = "num_participants")

# Plot the number of tests for each  length
test_lengths_healthy <- ggplot(test_length_counts, aes(x = test_length_rounded, y = num_participants)) +
  geom_bar(stat = "identity", fill = "skyblue", color = "black") +
  labs(
    title = "Distribution of Test Lengths (Rounded to Nearest Minute)",
    x = "Test Length (Rounded to Nearest Minute)",
    y = "Number of Participants"
  )
ggsave("test_lengths_healthy.png",plot = test_lengths_healthy, width = 10, height = 10)


## same for COPD
data_COPD <- data_COPD %>%
  group_by(PAN_ID) %>%
  mutate(test_length = max(Time, na.rm = TRUE)) %>%
  ungroup()

test_lengths <- data_COPD %>%
  group_by(PAN_ID) %>%
  summarise(test_length = max(Time, na.rm = TRUE)) %>%
  ungroup()

test_lengths <- test_lengths %>%
  mutate(test_length = as.numeric(test_length),
         test_length_rounded = round(test_length / 60))

test_length_counts <- test_lengths %>%
  count(test_length_rounded, name = "num_participants")

test_lengths_COPD <- ggplot(test_length_counts, aes(x = test_length_rounded, y = num_participants)) +
  geom_bar(stat = "identity", fill = "skyblue", color = "black") +
  labs(
    title = "Distribution of Test Lengths (Rounded to Nearest Minute)",
    x = "Test Length (Rounded to Nearest Minute)",
    y = "Number of Participants"
  )
ggsave("test_lengths_COPD.png",plot = test_lengths_COPD, width = 10, height = 10)


## same for asthma
data_asthma <- data_asthma %>%
  group_by(PAN_ID) %>%
  mutate(test_length = max(Time, na.rm = TRUE)) %>%
  ungroup()

test_lengths <- data_asthma %>%
  group_by(PAN_ID) %>%
  summarise(test_length = max(Time, na.rm = TRUE)) %>%
  ungroup()
test_lengths <- test_lengths %>%
  mutate(test_length = as.numeric(test_length),
         test_length_rounded = round(test_length / 60))

test_length_counts <- test_lengths %>%
  count(test_length_rounded, name = "num_participants")

test_lengths_asthma <- ggplot(test_length_counts, aes(x = test_length_rounded, y = num_participants)) +
  geom_bar(stat = "identity", fill = "skyblue", color = "black") +
  labs(
    title = "Distribution of Test Lengths (Rounded to Nearest Minute)",
    x = "Test Length (Rounded to Nearest Minute)",
    y = "Number of Participants"
  )
ggsave("test_lengths_asthma.png",plot = test_lengths_asthma, width = 10, height = 10)

```

