---
title: "data cleaning"
output: html_document
date: "2024-10-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# packages required
library(readr)
library(dplyr)
library(lubridate)
library(tidyr)
library(stringr)
library(dplyr)
library(zoo)
library(table1)
```

```{r import data}
################################################################################
##################   Import CPET data   ########################################
################################################################################
folder_path <- "PANAMA_CPET_files_download"

# importing all the CPET files from the folder

files <- list.files(path = folder_path, pattern = "\\.TXT$", full.names = TRUE)
all_data <- list()
pan_id_counter <- list()

# Looping through each file
for (file in files) {
  file_name <- basename(file)
  pan_id <- sub("^(PAN_\\d{4}).*", "\\1", file_name)
  data <- read_delim(file, delim = "!", escape_double = FALSE,  locale = locale(decimal_mark = ","), trim_ws = TRUE)
  
  # Account for multiple tests for 1 participant
  if (!is.null(pan_id_counter[[pan_id]])) {
    pan_id_counter[[pan_id]] <- pan_id_counter[[pan_id]] + 1
  } else {
    pan_id_counter[[pan_id]] <- 1
  }
  data$PAN_ID <- paste0(pan_id, "_", pan_id_counter[[pan_id]])
  
  all_data[[file_name]] <- data
}

cpet_data <- do.call(rbind, all_data)

# Rename variables with spaces
cpet_data <- cpet_data %>%
  rename(bp_sys = "BP sys", bp_dia = "BP dia")

# Remove rows where time is NA
cpet_data <- cpet_data %>%
  filter(!is.na(Time))

# Reformat time variable 
seconds <- as.numeric(cpet_data$Time/60)
cpet_data$Time <- hms::as_hms(seconds)

cpet_data$PAN_ID_short <- str_extract(cpet_data$PAN_ID, "^PAN_\\d{4}")

# Keep variables of interest to have a more manageable dataset
cpet_data <- cpet_data[,c("PAN_ID","PAN_ID_short","Time","Step","Load","VE","HR","RR","VO2","VCO2")]

# # Export, to have the csv at hand
# write.csv(cpet_data, file = "cpet_data.csv", row.names = FALSE)
# # Import file
cpet_data <- read_csv("cpet_data.csv")

################################################################################
################   Import REDCap data (descriptives)  ##########################
################################################################################

redcap_data <- read_csv("panama_redcap.csv")

## Join the clinical and descriptives into one row, no need to have them separated - one row per participant
redcap_data <- redcap_data %>%
  group_by(record_id) %>%
  summarise(across(everything(), ~ first(na.omit(.)), .names = "{col}")) %>%
  ungroup()

redcap_data <- redcap_data %>%
  rename(PAN_ID = cpet_id)
redcap_data$PAN_ID <- as.character(redcap_data$PAN_ID)

# add PAN_ID of participant with empty ID field (PAN_0141)
redcap_data <- redcap_data %>%
  mutate(PAN_ID = ifelse(record_id == 141, "PAN_0141", PAN_ID))

# Correct structure of the variables
redcap_data <- redcap_data %>%
  mutate(
    ethnic = as.factor(ethnic),
    smoke = as.factor(smoke),
    sex = as.factor(sex),
    disease = as.factor(disease),
    cpet_protocol = as.factor(cpet_protocol),
    copd_sev = as.factor(copd_sev),
    asthma_sev = as.factor(asthma_sev)
  )

# Rename IDs with spelling errors
redcap_data$PAN_ID[redcap_data$PAN_ID=='pAN_0043'] <- 'PAN_0043'
redcap_data$PAN_ID[redcap_data$PAN_ID=='PAN_110'] <- 'PAN_0110'
redcap_data$PAN_ID[redcap_data$PAN_ID=='PAN_154'] <- 'PAN_0154'
redcap_data$PAN_ID[redcap_data$PAN_ID=='PAN_159'] <- 'PAN_0159'
redcap_data$PAN_ID[redcap_data$PAN_ID=='PAN_161'] <- 'PAN_0161'
redcap_data$PAN_ID[redcap_data$PAN_ID=='PAN_162'] <- 'PAN_0162'
redcap_data$PAN_ID[redcap_data$PAN_ID=='PAN_164'] <- 'PAN_0164'
redcap_data$PAN_ID[redcap_data$PAN_ID=='PAN_172'] <- 'PAN_0172'


# Delete unnecessary variables - need to rewrite
# Keep variables of interest to have a more manageable dataset
redcap_data <- redcap_data[,c("PAN_ID","sex","ethnic","age","height","weight","smoke","freq_smoke",
                              "unit_smoke","mwt","mwt_dist","disease","fvc","fvc_pred","fev1","fev1_pred",
                              "fev1_fvc_r","tlc","tlcpred","rv","rvpred","rv_tlc","raw","sgaw",
                              "cpet_protocol","copd_sev","asthma_sev")]

# # save to have csv in folder
# write.csv(redcap_data,"redcap_data.csv",row.names = FALSE)
# # import
redcap_data <- read.csv("redcap_data.csv")

# join descriptives and CPET data
PAN_data <- cpet_data %>%
  left_join(redcap_data, by = c("PAN_ID_short" = "PAN_ID"))

PAN_data$PAN_ID <- as.factor(PAN_data$PAN_ID)
```

CLEAN DATA

Remove recovery time (as this could affect the prediction)
```{r}
PAN_data <- PAN_data %>%
  filter(!(Step %in% c("Rep", "R 2", "R 1")))
```


Missings
```{r}
# Table with NAs in each variable
missing_summary <- PAN_data %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Missing_Count")

print(missing_summary)
```


GENERATE VARIABLES
```{r}
# BMI (weight in kg/ height in m2)
PAN_data$bmi <- PAN_data$weight / ((PAN_data$height/100)^2)

# METs
PAN_data$METs <- ((PAN_data$VO2 * 1000) / PAN_data$weight ) / 3.5

# Respiratory Exchange Ratio
PAN_data$RER <- PAN_data$VCO2 / PAN_data$VO2

# VE max
PAN_data <- PAN_data %>%
  group_by(PAN_ID) %>%
  mutate(VE_max = max(VE, na.rm = TRUE)) %>%
  ungroup()

# RR max
PAN_data <- PAN_data %>%
  group_by(PAN_ID) %>%
  mutate(RR_max = max(RR, na.rm = TRUE)) %>%
  ungroup()

# HR max
PAN_data <- PAN_data %>%
  group_by(PAN_ID) %>%
  mutate(HR_max = max(HR, na.rm = TRUE)) %>%
  ungroup()

# VO2 max
PAN_data <- PAN_data %>%
  group_by(PAN_ID) %>%
  mutate(VO2_max = max(VO2, na.rm = TRUE)) %>%
  ungroup()

# exercise intensity
PAN_data <- PAN_data %>%
    mutate(
      ex_intensity = if_else(
        disease %in% c("Healthy", "Asthma"),
        case_when(
        # Age 20-39
        age <= 39 & METs < 2.4 ~ "Very Light",
        age <= 39 & METs < 4.8 ~ "Light",
        age <= 39 & METs < 7.2 ~ "Moderate",
        age <= 39 & METs < 10.2 ~ "Vigorous",
        age <= 39 & METs >= 10.2 ~ "Near to Maximal",
        
        # Age 40-64
        age >= 40 & age <= 64 & METs < 2.0 ~ "Very Light",
        age >= 40 & age <= 64 & METs < 4.0 ~ "Light",
        age >= 40 & age <= 64 & METs < 6.0 ~ "Moderate",
        age >= 40 & age <= 64 & METs < 8.5 ~ "Vigorous",
        age >= 40 & age <= 64 & METs >= 8.5 ~ "Near to Maximal",
        
        # Age >= 65
        age >= 65 & METs < 1.6 ~ "Very Light",
        age >= 65 & METs < 3.2 ~ "Light",
        age >= 65 & METs < 4.8 ~ "Moderate",
        age >= 65 & METs < 6.8 ~ "Vigorous",
        age >= 65 & METs >= 6.8 ~ "Near to Maximal",
        
        TRUE ~ NA_character_
      ),
              case_when(
          # Age 20-39
          age <= 39 & METs < 2.4 ~ "Very Light",
          age <= 39 & METs >= 2.4 & METs < 4.8 ~ "Light",
          age <= 39 & METs >= 4.8 & METs < 7.2 ~ "Moderate",
          age <= 39 & METs >= 7.2 & METs < 10.2 ~ "Vigorous",
          age <= 39 & METs >= 10.2 ~ "Near to Maximal",

          # Age 40-49
          age >= 40 & age <= 49 & METs < 2.0 ~ "Very Light",
          age >= 40 & age <= 49 & METs >= 2.0 & METs < 4.0 ~ "Light",
          age >= 40 & age <= 49 & METs >= 4.0 & METs < 6.0 ~ "Moderate",
          age >= 40 & age <= 49 & METs >= 6.0 & METs < 8.5 ~ "Vigorous",
          age >= 40 & age <= 49 & METs >= 8.5 ~ "Near to Maximal",

          # Age >= 50
          age >= 50 & METs < 1.6 ~ "Very Light",
          age >= 50 & METs >= 1.6 & METs < 3.2 ~ "Light",
          age >= 50 & METs >= 3.2 & METs < 4.8 ~ "Moderate",
          age >= 50 & METs >= 4.8 & METs < 6.8 ~ "Vigorous",
          age >= 50 & METs >= 6.8 ~ "Near to Maximal",

          TRUE ~ NA_character_
              )
      ),
      ex_intensity = factor(
        ex_intensity,
        levels = c("Very Light", "Light", "Moderate", "Vigorous", "Near to Maximal"),
        ordered = FALSE
      )
    )

# HR lag
PAN_data$Time <- hms::parse_hms(PAN_data$Time)

PAN_data <- PAN_data %>%
  mutate(Time_seconds = as.numeric(Time)) %>%
  arrange(Time_seconds)
PAN_data <- PAN_data %>%
  group_by(PAN_ID) %>%
  mutate(HR_lag = sapply(Time_seconds, function(t) {
    mean(HR[Time_seconds >= (t - 10) & Time_seconds < t], na.rm = TRUE)
  })) %>%
  ungroup()
PAN_data <- PAN_data %>%
  mutate(HR_lag = ifelse(is.na(HR_lag), HR, HR_lag))

# RR lag
PAN_data <- PAN_data %>%
  group_by(PAN_ID) %>%
  mutate(RR_lag = sapply(Time_seconds, function(t) {
    mean(RR[Time_seconds >= (t - 10) & Time_seconds < t], na.rm = TRUE)
  })) %>%
  ungroup()
PAN_data <- PAN_data %>%
  mutate(RR_lag = ifelse(is.na(RR_lag), RR, RR_lag))

# exercise intensity lag
PAN_data <- PAN_data %>%
  group_by(PAN_ID) %>%  # Group by participant ID
  mutate(EI_lag = ex_intensity[match(Time_seconds - 5, Time_seconds)]) %>% 
  ungroup() %>%
  mutate(EI_lag = ifelse(is.na(EI_lag), ex_intensity, EI_lag))

# Remove seconds variable generated
PAN_data <- PAN_data %>%
  dplyr::select(-Time_seconds)

```

Remove outliers (by participant) for time series variables (VE, HR and RR)
1. Group by participant
2. calculate moving average
3. calculate moving sd
4. identify values that are outside of 3 SDs from the mean
5. remove them

```{r remove outliers}
window_size <- 30 #have tried other values but 30 is the most backed by the literature

# minute ventilation 
PAN_data <- PAN_data %>%
  group_by(PAN_ID) %>%  # calculate mean and sd by participant
  mutate(
    MA = rollapply(VE, width = window_size, FUN = mean, align = "center", fill = NA),
    SD = rollapply(VE, width = window_size, FUN = sd, align = "center", fill = NA)
  ) %>%
  mutate(
    outlier = abs(VE - MA) > 3 * SD # detect datapoints outside of 3 SDs from the local mean
  ) %>%
  ungroup() %>% 
  dplyr::select(-c(MA,SD))

# Heart Rate - Process only if not an outlier
PAN_data <- PAN_data %>%
  group_by(PAN_ID) %>%
  mutate(
    # Only process non-outliers
    HR = ifelse(outlier, HR, HR),  # Keep HR unchanged for outliers
    MA = ifelse(!outlier, rollapply(HR, width = window_size, FUN = mean, align = "center", fill = NA), NA),
    SD = ifelse(!outlier, rollapply(HR, width = window_size, FUN = sd, align = "center", fill = NA), NA),
    outlier = ifelse(!outlier, abs(HR - MA) > 3 * SD, outlier)  # Update outliers only for non-marked rows
  ) %>%
  ungroup() %>%
  dplyr::select(-c(MA, SD))  # Remove temporary variables

# Breathing Rate - Process only if not an outlier, then filter
PAN_data <- PAN_data %>%
  group_by(PAN_ID) %>%
  mutate(
    # Only process non-outliers
    RR = ifelse(outlier, RR, RR),  # Keep RR unchanged for outliers
    MA = ifelse(!outlier, rollapply(RR, width = window_size, FUN = mean, align = "center", fill = NA), NA),
    SD = ifelse(!outlier, rollapply(RR, width = window_size, FUN = sd, align = "center", fill = NA), NA),
    outlier = ifelse(!outlier, abs(RR - MA) > 3 * SD, outlier)  # Update outliers only for non-marked rows
  ) %>%
  filter(!outlier) %>%  # Filter out rows marked as outliers
  ungroup() %>%
  dplyr::select(-c(MA, SD, outlier))  # Remove temporary variables

```

delete datapoints past peak VE, as they indicate a recovery phase that might not have been recorded as such
```{r}
# data_healthy <- data_healthy %>%
#   group_by(PAN_ID) %>%
#   mutate(max_index = which(VE == max(VE))[1]) %>%
#   filter(row_number() <= max_index) %>%
#   select(-max_index) %>%
#   ungroup()

```

Delete data points past anaerobic threshold (RER >= 1.0 consistently for 15 seconds)
```{r}
anaerobic_threshold <- PAN_data %>%
  group_by(PAN_ID) %>%
  arrange(Time) %>%
  mutate(RER_above_1 = RER >= 1.0,  # Identify where RER > 1.0
         anaerobic_threshold = cumsum(!RER_above_1),# Reset when RER <= 1.0
         flag = ave(RER_above_1, anaerobic_threshold, 
                    FUN = cumsum) > 30) %>%
  ungroup() %>%
  filter(flag)

PAN_data <- PAN_data %>%
  anti_join(anaerobic_threshold, by = c("PAN_ID", "Time"))
```


Disease as a factor variable with labels
```{r}
PAN_data <- PAN_data %>%
  mutate(
    disease = factor(
      disease,
      levels = c(1, 2, 3),
      labels = c("Healthy", "COPD", "Asthma")
    )
  )
```

Separate datasets by cohorts for exercise intensity variable generation 
```{r datasets by cohorts}
data_healthy <- PAN_data[PAN_data$disease == "Healthy",]
data_healthy <- data_healthy %>% dplyr::select(-c(copd_sev,asthma_sev))
data_COPD <- PAN_data[PAN_data$disease == "COPD",]
data_COPD <- data_COPD %>% dplyr::select(-asthma_sev)
data_asthma <- PAN_data[PAN_data$disease == "Asthma",]
data_asthma <- data_asthma %>% dplyr::select(-copd_sev)
```


REMOVE
Exercise intensity variable generation (by cohorts to adjust for disease)
```{r}
# Healthy and asthma - generate a function to not repeat twice
calculate_ex_intensity <- function(data) {
  data %>%
    mutate(
      ex_intensity = case_when(
        # Age 20-39
        age <= 39 & METs < 2.4 ~ "Very Light",
        age <= 39 & METs < 4.8 ~ "Light",
        age <= 39 & METs < 7.2 ~ "Moderate",
        age <= 39 & METs < 10.2 ~ "Vigorous",
        age <= 39 & METs >= 10.2 ~ "Near to Maximal",
        
        # Age 40-64
        age >= 40 & age <= 64 & METs < 2.0 ~ "Very Light",
        age >= 40 & age <= 64 & METs < 4.0 ~ "Light",
        age >= 40 & age <= 64 & METs < 6.0 ~ "Moderate",
        age >= 40 & age <= 64 & METs < 8.5 ~ "Vigorous",
        age >= 40 & age <= 64 & METs >= 8.5 ~ "Near to Maximal",
        
        # Age >= 65
        age >= 65 & METs < 1.6 ~ "Very Light",
        age >= 65 & METs < 3.2 ~ "Light",
        age >= 65 & METs < 4.8 ~ "Moderate",
        age >= 65 & METs < 6.8 ~ "Vigorous",
        age >= 65 & METs >= 6.8 ~ "Near to Maximal",
        
        TRUE ~ NA_character_
      ),
      ex_intensity = factor(
        ex_intensity,
        levels = c("Very Light", "Light", "Moderate", "Vigorous", "Near to Maximal"),
        ordered = FALSE
      )
    )
}

# Apply the function to healthy and asthma
data_healthy <- calculate_ex_intensity(data_healthy)
data_asthma <- calculate_ex_intensity(data_asthma)

## COPD - ages adjusted according to ACSM for disease
data_COPD <- data_COPD %>%
  mutate(
    ex_intensity = case_when(
      # Age 20-39
      age <= 39 & METs < 2.4 ~ "Very Light",
      age <= 39 & METs >= 2.4 & METs < 4.8 ~ "Light",
      age <= 39 & METs >= 4.8 & METs < 7.2 ~ "Moderate",
      age <= 39 & METs >= 7.2 & METs < 10.2 ~ "Vigorous",
      age <= 39 & METs >= 10.2 ~ "Near to Maximal",
      
      # Age 40-49
      age >= 40 & age <= 49 & METs < 2.0 ~ "Very Light",
      age >= 40 & age <= 49 & METs >= 2.0 & METs < 4.0 ~ "Light",
      age >= 40 & age <= 49 & METs >= 4.0 & METs < 6.0 ~ "Moderate",
      age >= 40 & age <= 49 & METs >= 6.0 & METs < 8.5 ~ "Vigorous",
      age >= 40 & age <= 49 & METs >= 8.5 ~ "Near to Maximal",
      
      # Age >= 50 (COPD patients over 50 are considered older adults (ACSM))
      age >= 50 & METs < 1.6 ~ "Very Light",
      age >= 50 & METs >= 1.6 & METs < 3.2 ~ "Light",
      age >= 50 & METs >= 3.2 & METs < 4.8 ~ "Moderate",
      age >= 50 & METs >= 4.8 & METs < 6.8 ~ "Vigorous",
      age >= 50 & METs >= 6.8 ~ "Near to Maximal",
      
      TRUE ~ NA_character_
    ),
    ex_intensity = factor(
      ex_intensity,
      levels = c("Very Light", "Light", "Moderate", "Vigorous", "Near to Maximal"),
      ordered = TRUE
    )
  )

```


exportar datasets
```{r}
write.csv(data_healthy, file = "data_healthy_08_01.csv", row.names = FALSE)
write.csv(data_COPD, file = "data_COPD_18_01.csv", row.names = FALSE)
write.csv(data_asthma, file = "data_asthma_08_01.csv", row.names = FALSE)
write.csv(PAN_data, file = "PAN_data.csv", row.names = FALSE)

```


EXPLORATORY DATA ANALYSIS
using Datamaid
```{r}
library(dataMaid)
# generate reports
makeDataReport(PAN_data, output = "html", replace = TRUE)
makeDataReport(data_COPD, output = "html", replace = TRUE)
makeDataReport(data_healthy, output = "html", replace = TRUE)
makeDataReport(data_asthma, output = "html", replace = TRUE)

```


DESCRIPTIVES to generate descriptive statistics tables

```{r}
### PARTICIPANT DESCRIPTORS
table1(~ factor(sex) + age + height + weight + factor(ethnic) +  factor(smoke) + factor(asthma_sev) + factor(copd_sev) + fev1 + fev1_pred + fvc + fvc_pred + fev1_fvc_r | disease, data=redcap_data, overall=c(left="Total"))

# treatment table
table1(~ factor(treatment___1) + factor(treatment___2) + factor(treatment___3) + factor(treatment___4) + factor(treatment___5) + factor(treatment___6) + factor(treatment___7) + factor(treatment___8)| disease, data=redcap_data, overall=c(left="Total"))

```


Descriptives of physiological variables

```{r}
## PHYSIOLOGICAL VARIABLES DESCRIPTORS
table1(~ VE + Ti + RR + VO2 + HR | disease, data=PAN_data, overall=c(left="Total"))

# Rest is when step is Rep (will only include some of the participants as they dont all have rep)
data_rest <- PAN_data[(PAN_data$Step=='Rep'),]

# values at rest
table1(~ VE + Ti + RR + VO2 + HR | disease, data=data_rest, overall=c(left="Total"))
```


Maximum values' descriptives
```{r}
### VE max mean(SD)
# Total
summary_stats <- PAN_data %>%
  group_by(PAN_ID) %>%
  summarise(VE_max = unique(VE_max)) %>% # Take unique VE_max per participant
  summarise(mean_VE_max = mean(VE_max, na.rm = TRUE),
            sd_VE_max = sd(VE_max, na.rm = TRUE))
summary_stats
# No RD
summary_stats <- data_healthy %>%
  group_by(PAN_ID) %>%
  summarise(VE_max = unique(VE_max)) %>% # Take unique VE_max per participant
  summarise(mean_VE_max = mean(VE_max, na.rm = TRUE),
            sd_VE_max = sd(VE_max, na.rm = TRUE))
summary_stats
# COPD
summary_stats <- data_COPD %>%
  group_by(PAN_ID) %>%
  summarise(VE_max = unique(VE_max)) %>% # Take unique VE_max per participant
  summarise(mean_VE_max = mean(VE_max, na.rm = TRUE),
            sd_VE_max = sd(VE_max, na.rm = TRUE))
summary_stats
# Asthma
summary_stats <- data_asthma %>%
  group_by(PAN_ID) %>%
  summarise(VE_max = unique(VE_max)) %>% # Take unique VE_max per participant
  summarise(mean_VE_max = mean(VE_max, na.rm = TRUE),
            sd_VE_max = sd(VE_max, na.rm = TRUE))
summary_stats

### RR max mean(SD)
# Total
summary_stats <- PAN_data %>%
  group_by(PAN_ID) %>%
  summarise(RR_max = unique(RR_max)) %>% # Take unique RR_max per participant
  summarise(mean_RR_max = mean(RR_max, na.rm = TRUE),
            sd_RR_max = sd(RR_max, na.rm = TRUE))
summary_stats
# No RD
summary_stats <- data_healthy %>%
  group_by(PAN_ID) %>%
  summarise(RR_max = unique(RR_max)) %>% # Take unique RR_max per participant
  summarise(mean_RR_max = mean(RR_max, na.rm = TRUE),
            sd_RR_max = sd(RR_max, na.rm = TRUE))
summary_stats
# COPD
summary_stats <- data_COPD %>%
  group_by(PAN_ID) %>%
  summarise(RR_max = unique(RR_max)) %>% # Take unique RR_max per participant
  summarise(mean_RR_max = mean(RR_max, na.rm = TRUE),
            sd_RR_max = sd(RR_max, na.rm = TRUE))
summary_stats
# Asthma
summary_stats <- data_asthma %>%
  group_by(PAN_ID) %>%
  summarise(RR_max = unique(RR_max)) %>% # Take unique RR_max per participant
  summarise(mean_RR_max = mean(RR_max, na.rm = TRUE),
            sd_RR_max = sd(RR_max, na.rm = TRUE))
summary_stats

### HR max mean(SD)
# Total
summary_stats <- PAN_data %>%
  group_by(PAN_ID) %>%
  summarise(HR_max = unique(HR_max)) %>% # Take unique HR_max per participant
  summarise(mean_HR_max = mean(HR_max, na.rm = TRUE),
            sd_HR_max = sd(HR_max, na.rm = TRUE))
summary_stats
# No RD
summary_stats <- data_healthy %>%
  group_by(PAN_ID) %>%
  summarise(HR_max = unique(HR_max)) %>% # Take unique HR_max per participant
  summarise(mean_HR_max = mean(HR_max, na.rm = TRUE),
            sd_HR_max = sd(HR_max, na.rm = TRUE))
summary_stats
# COPD
summary_stats <- data_COPD %>%
  group_by(PAN_ID) %>%
  summarise(HR_max = unique(HR_max)) %>% # Take unique HR_max per participant
  summarise(mean_HR_max = mean(HR_max, na.rm = TRUE),
            sd_HR_max = sd(HR_max, na.rm = TRUE))
summary_stats
# Asthma
summary_stats <- data_asthma %>%
  group_by(PAN_ID) %>%
  summarise(HR_max = unique(HR_max)) %>% # Take unique HR_max per participant
  summarise(mean_HR_max = mean(HR_max, na.rm = TRUE),
            sd_HR_max = sd(HR_max, na.rm = TRUE))
summary_stats

### VO2 max mean(SD)
# Total
summary_stats <- PAN_data %>%
  group_by(PAN_ID) %>%
  summarise(VO2_max = unique(VO2_max)) %>% # Take unique VO2_max per participant
  summarise(mean_VO2_max = mean(VO2_max, na.rm = TRUE),
            sd_VO2_max = sd(VO2_max, na.rm = TRUE))
summary_stats
# No RD
summary_stats <- data_healthy %>%
  group_by(PAN_ID) %>%
  summarise(VO2_max = unique(VO2_max)) %>% # Take unique VO2_max per participant
  summarise(mean_VO2_max = mean(VO2_max, na.rm = TRUE),
            sd_VO2_max = sd(VO2_max, na.rm = TRUE))
summary_stats
# COPD
summary_stats <- data_COPD %>%
  group_by(PAN_ID) %>%
  summarise(VO2_max = unique(VO2_max)) %>% # Take unique VO2_max per participant
  summarise(mean_VO2_max = mean(VO2_max, na.rm = TRUE),
            sd_VO2_max = sd(VO2_max, na.rm = TRUE))
summary_stats
# Asthma
summary_stats <- data_asthma %>%
  group_by(PAN_ID) %>%
  summarise(VO2_max = unique(VO2_max)) %>% # Take unique VO2_max per participant
  summarise(mean_VO2_max = mean(VO2_max, na.rm = TRUE),
            sd_VO2_max = sd(VO2_max, na.rm = TRUE))
summary_stats
```


lo de arriba optimizado
```{r}
library(dplyr)

# Function to calculate mean and sd of a given variable
calculate_summary <- function(data, group_var, measure_var) {
  data %>%
    group_by({{ group_var }}) %>%
    summarise(!!measure_var := unique(.data[[measure_var]])) %>% # Take unique measure_var per group
    summarise(
      mean = mean(.data[[measure_var]], na.rm = TRUE),
      sd = sd(.data[[measure_var]], na.rm = TRUE)
    )
}

# Datasets list
datasets <- list(
  Total = PAN_data,
  No_RD = data_healthy,
  COPD = data_COPD,
  Asthma = data_asthma
)

# Variables to summarize
variables <- c("VE_max", "RR_max", "HR_max", "VO2_max")

# Loop through datasets and variables to calculate statistics
results <- list()

for (var in variables) {
  for (name in names(datasets)) {
    dataset <- datasets[[name]]
    stats <- calculate_summary(dataset, PAN_ID, var)
    results[[paste(name, var, sep = "_")]] <- stats
  }
}

# Print results
results

```


Length of CPETs
```{r}
# healthy cohort
test_lengths <- data_healthy %>%
  group_by(PAN_ID) %>%
  summarise(test_length = max(Time, na.rm = TRUE)) %>%
  ungroup()

test_lengths <- test_lengths %>%
  mutate(test_length = as.numeric(test_length),
         test_length_rounded = round(test_length / 60))

test_length_counts <- test_lengths %>%
  count(test_length_rounded, name = "num_participants")

# Plot the number of tests for each length
test_lengths_healthy <- ggplot(test_length_counts, aes(x = test_length_rounded, y = num_participants)) +
  geom_bar(stat = "identity", fill = "skyblue", color = "black") +
  labs(
    title = "Distribution of Test Lengths (Rounded to Nearest Minute)",
    x = "Test Length (Rounded to Nearest Minute)",
    y = "Number of Participants"
  )
ggsave("test_lengths_healthy.png",plot = test_lengths_healthy, width = 10, height = 10)


# same for COPD
test_lengths <- data_COPD %>%
  group_by(PAN_ID) %>%
  summarise(test_length = max(Time, na.rm = TRUE)) %>%
  ungroup()

test_lengths <- test_lengths %>%
  mutate(test_length = as.numeric(test_length),
         test_length_rounded = round(test_length / 60))

test_length_counts <- test_lengths %>%
  count(test_length_rounded, name = "num_participants")

test_lengths_COPD <- ggplot(test_length_counts, aes(x = test_length_rounded, y = num_participants)) +
  geom_bar(stat = "identity", fill = "skyblue", color = "black") +
  labs(
    title = "Distribution of Test Lengths (Rounded to Nearest Minute)",
    x = "Test Length (Rounded to Nearest Minute)",
    y = "Number of Participants"
  )
ggsave("test_lengths_COPD.png",plot = test_lengths_COPD, width = 10, height = 10)


# same for asthma
test_lengths <- data_asthma %>%
  group_by(PAN_ID) %>%
  summarise(test_length = max(Time, na.rm = TRUE)) %>%
  ungroup()
test_lengths <- test_lengths %>%
  mutate(test_length = as.numeric(test_length),
         test_length_rounded = round(test_length / 60))

test_length_counts <- test_lengths %>%
  count(test_length_rounded, name = "num_participants")

test_lengths_asthma <- ggplot(test_length_counts, aes(x = test_length_rounded, y = num_participants)) +
  geom_bar(stat = "identity", fill = "skyblue", color = "black") +
  labs(
    title = "Distribution of Test Lengths (Rounded to Nearest Minute)",
    x = "Test Length (Rounded to Nearest Minute)",
    y = "Number of Participants"
  )
ggsave("test_lengths_asthma.png",plot = test_lengths_asthma, width = 10, height = 10)

```

TIME RESOLUTIONS - 5s, 15s, 30s, 1min

```{r}
# currently at breath by breath

# Time in POSIXct format
cpet_data$Time <- as.POSIXct(cpet_data$Time, format = )

###### 5 second intervals
cpet_data_10s <- cpet_data %>%
  mutate(Time = floor_date(Time, unit = "1 second")) %>% 
  group_by(PAN_ID, Time_10s) %>%
  summarise(across(everything(), mean, na.rm = TRUE), .groups = 'drop')

cpet_data_10s$Time_10s <- as.numeric(cpet_data_10s$Time_10s)
# Fix Time variable
seconds <- as.numeric(cpet_data_10s$Time_10s)
cpet_data_10s$Time_10s <- hms::as_hms(seconds)
cpet_data_10s$Time <- cpet_data_10s$Time_10s


######  30 second intervals
cpet_data_30s <- cpet_data %>%
  mutate(Time_30s = floor_date(Time, unit = "30 seconds")) %>%
  group_by(PAN_ID, Time_30s) %>%
  summarise(across(everything(), mean, na.rm = TRUE), .groups = 'drop')

cpet_data_30s$Time_30s <- as.numeric(cpet_data_30s$Time_30s)
# Fix Time variable
seconds <- as.numeric(cpet_data_30s$Time_30s)
cpet_data_30s$Time_30s <- hms::as_hms(seconds)
cpet_data_30s$Time <- cpet_data_30s$Time_30s


######  1 minute intervals
cpet_data_60s <- cpet_data %>%
  mutate(Time_60s = floor_date(Time, unit = "1 minute")) %>%
  group_by(PAN_ID, Time_60s) %>%
  summarise(across(everything(), mean, na.rm = TRUE), .groups = 'drop')

cpet_data_60s$Time_60s <- as.numeric(cpet_data_60s$Time_60s)
# Fix Time variable
seconds <- as.numeric(cpet_data_60s$Time_60s)
cpet_data_60s$Time_60s <- hms::as_hms(seconds)
cpet_data_60s$Time <- cpet_data_60s$Time_60s


# merge redcap data with aggregated CPET data (for )
PAN_data_10s <- cpet_data_10s %>%
  left_join(redcap_data, by =  c("PAN_ID_short" = "PAN_ID"))

PAN_data_30s <- cpet_data_30s %>%
  left_join(redcap_data, by =  c("PAN_ID_short" = "PAN_ID"))

PAN_data_60s <- cpet_data_60s %>%
  left_join(redcap_data, by = c("PAN_ID_short" = "PAN_ID"))

```


